Optimal Neural Summarisation for Full-Field Implicit Inference by Density Estimation

### 2023-07-24 paper/paper.tex
Updates from Overleaf

   \justine{\textbf{I think the 2 main ideas for this part are: }

    - classical way to overcome this issue: we compress the data into 2pt statistics  and use gaussian likelihood → pb: we miss a lot of information}
    \denise{That's true. But adding it in the abstract would make the "Context" subsection (this division is mandatory for A$\&$A) too long. This is properly described in the introduction}

    \justine{ - considering the previous pb: thanks to Implicit Inference methods we can consider any kind of summary stat (because we can learn the exact posterior p(theta | summary stat)), so to have the best contours (meaning that we exatrct all gaussian as well as non-gaussian signal) we only have to focus on how we compress our data.}
    \denise{That's also true. But we don't use any summary statistics (except the power spectrum because it's easy), and I think that one of the points of the paper is to extract the information from the pixels, without using any summaries. So I personally would not talk about summary statistics }

     \justine{\textbf{Note:}

    - for me LFI = Implicit Inference, I think we should use only one
   }
   \denise{True again. But the definition "Likelihood-Free Inference" is more used in the community. Implicit Inference is something that we introduce in the motivation and introduction. I don't know exactly how much people are familiar with this concept. }
    \justine{ - 'Likelihood-Free Inference (LFI) approaches offer a promising solution by using forward-simulated synthetic data instead of an explicit likelihood function.' → Implicit inference approaches offer a promising solution to approximate posterior distribution by using simulations from an implicit likelihood (a simulator) instead of evaluating explicit likelihood function ..?

    - I'm probably wrong but not sure about the last sentence.
}
   }   
  % aims heading (mandatory)
    {We aim to derive the optimal neural-compression strategy for full field Likelihood free inference by density estimation. Moreover, by using the optimal compression strategy, we demonstrate that the posterior distribution is comparable to those derived from a Bayesian forward modeling approach.}  
  % methods heading (mandatory)
    { We present a comparative analysis of different loss functions employed during the training of the neural network. We  evaluate their performance by measuring their impact on the constraints on the $\Lambda$CDM parameters expected from LSST-Y10.
    For both forward-modeling the convergence field and generating observed mock data, we use the \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\url{SbiLens}} package. 
    \texttt{SbiLens} is a jax-based differentiable physical model specifically designed for Likelihood Free Inference and Bayesian inference algorithms that require access to likelihood derivatives.

    \justine{\textbf{I think the main ideas for this part are: }

    - first we describe sbi\_lens : Provides a differentiable log normal mass map simulator specifically designed for for weak lensing inference. Then you can describe a bit the simulations like tomographic bins etc I guess

    - We construct our ground truth

    - We use implicit inference techniques to go from the compressed stat to the posteriors}
    }
  % results heading (mandatory)
   {Our experiments validate the effectiveness of LFI methods,  as we demonstrate comparable posterior distributions between LFI and Bayesian forward modeling approaches \justine{field-based Bayesian Hierarchical Model analysis}. Additionally, our results provide insights into the design of data compression procedures to derive low-dimensional summary statistics while minimizing information loss. All codes associated with this paper can be found at this link \github.}
  % conclusions heading (optional), leave it empty if necessary 
   {}
   \keywords{methods: statistical – gravitational lensing: weak – cosmology: large-scale structure of Universe
               }

@@

%-------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{Introduction}
\denise{PEOPLE CAN READ THIS!!}
Weak gravitational lensing by the large-scale structure is caused by the presence of foreground matter bending the light emitted by background galaxies. Being sensitive to the large-scale structure of the universe, it is one of the most promising tools for investigating the nature of dark energy, the origin of the accelerating expansion of the universe and estimating cosmological parameters. Future cosmological surveys, such as the Legacy Survey of Space and Time (LSST) of the Vera C. Rubin Observatory \citep{ivezic2019lsst}, the Nancy Grace Roman Space Telescope \citep{spergel2015wide}, and  the Euclid Mission \citep{laureijs2011euclid}, will rely on weak gravitational lensing as one of the principal physical probes to address unresolved questions in current cosmology.
 As these surveys become deeper, they will be able to access more non-Gaussian features of the matter fields. This makes the standard weak lensing analyses, which rely on two-point statistics such as the 2-point shear correlation or the angular power spectrum, sub-optimal. These analyses are unable to fully capture the non-Gaussian information imprinted in the lensing signal and can only access Gaussian information.

@@

In this paper, we present a comparison of the performance of various neural network-based compression schemes within the Likelihood-Free Inference (LFI) framework. These methods differ in terms of the loss functions used to train the neural network, but they share the same inference strategy based on neural density estimation. To address the need for result validation in the LFI context, we employ a Bayesian forward modeling approach that enables inference on the joint posterior of the cosmological parameters and the convergence field. The forward model is based on \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\url{SbiLens}}, a Python package designed for weak lensing implicit inference with differentiable simulator implemented in Jax. \texttt{SbiLens} allows the sampling of convergence maps in a tomographic setting, accounting for the cross-correlation between different redshift bins.

The paper is structured as follows: in \autoref{Sec:Motivation}, we illustrate the motivation behind this work. In \autoref{Sec:the SBILens framework}, we introduce the \texttt{SbiLens} framework and describe the simulated data used in this work. In \autoref{Sec:experiment}, we detail the inference strategy and the three different approaches we used: the power spectrum, the map-based inference based on the Bayesian forward-modeling, and the map-based inference based on the LFI. In the same section, we also provide a detailed overview of different neural compression strategies. In \autoref{Sec:results}, we discuss the results and validate the LFI approaches. Finally, we conclude in \autoref{Sec:conclusion}. 
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{Motivation}\label{Sec:Motivation}
%--------------------------------------------------------------------
With the increased statistical power of stage IV surveys, our cosmological analysis should not rely on the measurement of sub-optimal summary statistics, that may not fully capture the non-Gaussian information present in the lensing field at the scales accessible to future surveys. In this paper we introduce a forward model that directly extracts information from the raw pixel data, rather than relying on the analytical evaluation of summary statistics. By doing so, we aim to preserve all available information and facilitate the incorporation of systematic effects and the combination of multiple cosmological probes through joint simulations.
In this context, the simulator of the observables serves as our physical model, where each component is tractable. These models, often refereed to as \textit{probabilistic program}, can be described as follows. They take as input a vector parameter $\bm{\theta}$, which describes the underlying deterministic model. Then, they sample internal states, dubbed \textit{latent variables}, from the distribution $\bm{z}_i \sim p_i(\bm{z}_i|\bm \theta, \bm{z}_{<i})$, than can be directly or indirectly related to a physically meaningful state of the system. Finally, they generate the output $\bm x$ from the distribution $\ p(\bm d|\bm \theta)$, where $\bm{d}$ represents the observations.

The ultimate goal of Bayesian inference in cosmology is to compute the posterior distribution:
\begin{equation}\label{Eq:posterior}
     p(\bm{\theta}|\bm{x})= 
     \frac{p(\bm{x}|\bm{\theta})p(\bm{\theta})}
     {\int d\bm{\theta'}p(\bm{x}|\bm{\theta}')p(\bm{\theta}')},
\end{equation}
however, a problems arises because the marginal likelihood $p(\bm{x}|\bm{\theta})$ is typically intractable:
\begin{equation}
    p(\bm{x}|\bm{\theta})=\int p(\bm{x},\bm{z}|\bm{\theta}) d\bm{z}=\int p(\bm{x}|\bm{z},\bm{\theta})p(\bm{z}|\bm{\theta}) d\bm{z}.
\end{equation}
as it corresponds to an integral over all possible trajectories throughout the latent space.
To overcome this limitation while still capturing the full information content of the data, two different approaches have been proposed in the literature. Although these approaches are often referred to by different names, in this work, we will make the following distinction:
%--------------------------------------------------------------------
\paragraph{\textbf{Explicit inference}} referring to all Likelihood-based inference approaches.  
This approach involves treating the simulator as a probabilistic model and performing inference over the joint posterior:
\begin{equation}
        p(\bm{\theta},\bm{z}|\bm{x})\propto  p(\bm{x},\bm{z}|\bm{\theta}) p(\bm{z}|\bm{\theta}).
\end{equation}
In the context of the forward model and full field analysis, fall in this category the Bayesian Hierarchical Models (BHMs).
A Bayesian forward approach involves using a given physical model to predict observations and then comparing these predictions with observations to infer the parameters of the model.
However, performing a global analysis on complex and large datasets, such as a lensing dataset, is not feasible. Therefore, the idea of dividing the global analysis into several subgroups, analyzing them in multiple steps, and using the output of each of those as input for the next one. From here, the \textit{hierarchical} nature of this methodology.
%--------------------------------------------------------------------
\paragraph{\textbf{Implicit inference}} referring to all the approaches not relying on an analytical model to describe the signal, but rather on learning the likelihood from simulations. This second class of approaches involves treating the simulator as a black box with only the ability to sample from the joint distribution:
\begin{equation}
    (\bm{x}, \bm{\theta})\sim p(\bm{x}, \bm{\theta}).
\end{equation}
Within this class of methods, we can differentiate between more traditional methods such as Approximate Bayesian Computation (ABC) and Density Estimation Likelihood-Free Inference (DELFI) methods. ABC employs rejection-criteria based approaches to approximate the likelihood by comparing simulations with data. 
In this work, our focus will be on DELFI methods, which approach the inference task as a density estimation problem. \\
We can divide Likelihood-Free Inference approaches into two separate steps:
\begin{enumerate}
    \item  Automatically learning an optimal low-dimensional summary statistic.
    \item Using Neural Density Estimation in low dimensions to infer the target distributions.
\end{enumerate}
In the first step, we introduce a parametric function $f_{\varphi}$ such that:
     \begin{equation}
         \bm{y}=f_{\varphi}(\bm{\theta}),
     \end{equation}
which aims to reduce the dimensionality of the data while preserving information. Typically, the compressed statistics $\bm{y}$ is assumed to have the same dimension of $\bm{\theta}$. \\
In the second step, Neural Density Estimation can target either building an estimate $p_{\phi}$ of the likelihood function $p(\bm{x}|\bm{\theta})$ (referred to as the Neural Likelihood Estimation (NLE) task), or targeting the posterior distribution $p(\bm{\theta}|\bm{x})$, (known as Neural Posterior Estimation (NPE) task).

@@

%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{The SBILens framework}\label{Sec:the SBILens framework}
\denise{PEOPLE CAN READ THIS!!}
\subsection{Lognormal Modeling}\label{Sec:Lognormal Modeling}
For various cosmological applications, the non-Gaussian field can be modeled as a Lognormal field \citep{coles1991lognormal,bohm2017bayesian}.
This model offers the advantage of generating the matter or convergence field rapidly while allowing the extraction of information beyond the two-point statistics. 
Although studies demonstrated that this model fails in describing the 3D field \citep{klypin2018density}, it properly describes the 2D convergence field \citep{clerkin2017testing, xavier2016improving}.
Assuming a simulated Gaussian convergence map $\kappa_g$, whose statistical properties are fully described by its power spectrum $C_{\ell}$ we know that this model is not a suitable representation of late-time and more evolved structures. One potential solution is to find a transformation $f(\kappa_g)$ of this map that captures the non-Gaussian features in the convergence field. In doing so, it is crucial to ensure that the transformed map maintains the correct mean and variance, effectively recovering the correct two-point statistics.
Denoting $\mu$ and $\sigma_g^2$ the mean and covariance matrix of $\kappa_g$ respectively, we can define the transformed convergence $\kappa_{ln}$ as a shifted lognormal random field:
\begin{equation}\label{Eq:log_norm_kappa}
    \kappa_{ln}=e^{\kappa_{g}}-\lambda, 
\end{equation}
where $\lambda$ is a free parameter that determines the shift of the lognormal distribution. The convergence $\kappa$ in a given redshift bin is fully determined by the shift parameter $\lambda$, the mean $\mu$ of the associated Gaussian field $\kappa_g$, and its variance $\sigma_{g}^2\equiv \xi_g$.
The correlation of the lognormal field, denoted as $\xi_{ln}$, is also a function of these variables and is related to $\xi^{ij}_g$ through the following equations:
\begin{align}
    \xi^{ij}_{ln}(\theta) & \equiv \lambda_i \lambda_j (e^{ \xi^{ij}_g(\theta)}-1) \nonumber \\ 
    \xi^{ij}_g(\theta)&=\log{\left[ \frac{\xi^{ij}_{ln}(\theta)}{\lambda_i \lambda_j}+1\right ]}. \label{Eq:log_norm_corr}
\end{align}
Here $i$ and $j$ define a pair of redshift bins.
The parameter $\lambda$, also known as \textit{minimum convergence
parameter}, defines the lowest values for all possible values of $\kappa$.
The modeling of the shift parameter can be approached in various ways. For example, it can be determined by matching moments of the distribution \citep{xavier2016improving} or by treating it as a free parameter \citep{hilbert2011cosmic}. In general, the value of $\lambda$ depends on the redshift, cosmology, and the scale of the field at which smoothing is applied.

@@

As mentioned by \citet{boruah2022map}, the perturbation theory-based approach may not provide accurate results at small scales. However, the main objective of this paper is to compare different inference strategies and we are not concerned with the potential implications of this approximation. 
In future applications of the presented inference pipeline, the simulation procedure will be based on N-body simulations.

@@

@@

For each redshift bin, we assume Gaussian noise with mean zero and variance given by

@@

@@

@@

\denise{PEOPLE CAN READ THIS!!}

@@

This means that this loss function is actually sub-optimal for two reasons:
\begin{itemize}
    \item The summary extracted by the neural network $t$ is, similarly to the MSE case only an estimate of the mean of 
    the posterior distribution, which is not guaranteed to be sufficient.
    \item Because of the Gaussian variational assumption, the variational posterior can potentially be biased with respect to the true posterior, and thus the mean of this Gaussian approximation may be biased with respect to the true posterior mean.
\end{itemize}
 A simple MSE would thus seem to be preferable.
}
%--------------------------------------------------------------------
\paragraph{\textcolor{violet}
{Information Maximising Neural Networks (IMNNs)}}. \LM{TODO: condensed introduction; emphasize learning compression at single point; network learns score function which generalises smoothly away from fiducial \cite{Makinen_2021, Makinen_2022, Charnock_2018}, Wandelt (in prep.) 
}

@@

%--------------------------------------------------------------------
%--------------------------------------------------------------------
%--------------------------------------------------------------------
\subsubsection{Inference strategy}
%--------------------------------------------------------------------
%--------------------------------------------------------------------
By comparing the posteriors $p(\theta| t)$ obtained with different compression strategies, we can assess the sensitivity of the results to the choice of statistics. In particular, the more informative the summary statistics, the “closer” the posterior $p(\theta | t)$ is to the true posterior $p(\theta | x)$. In this section, we describe how we approximate the posterior distribution for each summary statistic t, enable us to use it as a fair benchmark for comparison (we describe the performance metric in section ??).\\
Since the likelihood $p(t|\theta)$ is intractable, we propose to do Neural Posterior Estimation \citep{npe1, npe2, npe3}, a Simulation-Based Inference (SBI) strategy that aims to directly approximate, from simulations, the posterior distribution

\begin{equation}
    p(\theta | t) \propto \int p(t | \theta, z) p(z|\theta) p(\theta) dz\,.
\end{equation}

@@

For each summary statistic, we approximate it posterior distribution using the same NF architecture, namely a RealNVP \citep{realnvp} with 4 coupling layers, shift and scale parameters are learned using a neural network with 2 layers of 128 neurons, activation functions are $silu$ \citep{silu}. And, for each summary statistic, the prior $p(\theta)$ used in the inference process remains the same.
%--------------------------------------------------------------------
%--------------------------------------------------------------------
\subsection{Result estimators}
%--------------------------------------------------------------------
%--------------------------------------------------------------------
To evaluate the results obtained from different approaches, we use two commonly employed estimators found in the literature. Their definitions are provided in the following sections.
%--------------------------------------------------------------------
\subsubsection{Figure of Merit (FoM)}
%--------------------------------------------------------------------
To quantify the constraining power of each inference strategy, we first employ the Figure of Merit (FoM), defined as follows:
\begin{equation}\label{Eq:Figure_of_merit}
    \text{FoM}_{\alpha \beta}=\sqrt{\text{det}(\Tilde{F}_{\alpha \beta})}.
\end{equation}
Here, $\alpha$ and $\beta$ represent a pair of cosmological parameters, and $\Tilde{F}_{\alpha \beta}$ refers to the marginalized Fisher matrix. We calculate $\Tilde{F}_{\alpha \beta}$ as the inverse of the parameter space covariance matrix $C_{\alpha \beta}$, which is estimated from the HMC chains and from the Normalizing flow sampling. 
Under the assumption of a Gaussian covariance, the FoM defined in \autoref{Eq:Figure_of_merit} is proportional to the inverse of the $2-\sigma$ contours in the 2-dimensional marginalised parameter space of the $\alpha$ and $\beta$ pair.
%--------------------------------------------------------------------
\subsubsection{Kullback-Leibler divergence}
%--------------------------------------------------------------------
To measure the difference between the probability distribution obtained with different compressed summaries we compute the \textit{Kullback-Leibler divergence} or KL-divergence. As mentioned when we introduced the VMIM, the KL divergence is related to the relative entropy, and is a non-symmetric measure of the difference between two distributions. \denise{INCOMPLETE}
%--------------------------------------------------------------------
% ############# Begin Table BiblioSruvey  #############
%--------------------------------------------------------------------

>>>>

%    \justine{\textbf{I think the 2 main ideas for this part are: }

%     - classical way to overcome this issue: we compress the data into 2pt statistics  and use gaussian likelihood → pb: we miss a lot of information}
%     \denise{That's true. But adding it in the abstract would make the "Context" subsection (this division is mandatory for A$\&$A) too long. This is properly described in the introduction}

%     \justine{ - considering the previous pb: thanks to Implicit Inference methods we can consider any kind of summary stat (because we can learn the exact posterior p(theta | summary stat)), so to have the best contours (meaning that we exatrct all gaussian as well as non-gaussian signal) we only have to focus on how we compress our data.}
%     \denise{That's also true. But we don't use any summary statistics (except the power spectrum because it's easy), and I think that one of the points of the paper is to extract the information from the pixels, without using any summaries. So I personally would not talk about summary statistics }

%      \justine{\textbf{Note:}

%     - for me LFI = Implicit Inference, I think we should use only one
%    }
%    \denise{True again. But the definition "Likelihood-Free Inference" is more used in the community. Implicit Inference is something that we introduce in the motivation and introduction. I don't know exactly how much people are familiar with this concept. }
%     \justine{ - 'Likelihood-Free Inference (LFI) approaches offer a promising solution by using forward-simulated synthetic data instead of an explicit likelihood function.' → Implicit inference approaches offer a promising solution to approximate posterior distribution by using simulations from an implicit likelihood (a simulator) instead of evaluating explicit likelihood function ..?

%     - I'm probably wrong but not sure about the last sentence.
% }
   }   
  % aims heading (mandatory)
    {We aim to derive the optimal neural-compression strategy for full field Likelihood free inference by density estimation. Moreover, by using the optimal compression strategy, we demonstrate that the posterior distribution is comparable to those derived from a Bayesian forward modeling approach.}  
  % methods heading (mandatory)
    { We present a comparative analysis of different loss functions employed during the training of the neural network. We  evaluate their performance by measuring their impact on the constraints on the $\Lambda$CDM parameters expected from LSST-Y10.
    For both forward-modeling the convergence field and generating observed mock data, we use the \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\url{SbiLens}} package. 
    \texttt{SbiLens} is a jax-based differentiable physical model specifically designed for Likelihood Free Inference and Bayesian inference algorithms that require access to likelihood derivatives.

    % \justine{\textbf{I think the main ideas for this part are: }

    % - first we describe sbi\_lens : Provides a differentiable log normal mass map simulator specifically designed for for weak lensing inference. Then you can describe a bit the simulations like tomographic bins etc I guess

    % - We construct our ground truth

    % - We use implicit inference techniques to go from the compressed stat to the posteriors}
    }
  % results heading (mandatory)
   {Our experiments validate the effectiveness of LFI methods,  as we demonstrate comparable posterior distributions between LFI and Bayesian forward modeling approaches.
   % \justine{field-based Bayesian Hierarchical Model analysis}. 
   Additionally, our results provide insights into the design of data compression procedures to derive low-dimensional summary statistics while minimizing information loss. All codes associated with this paper can be found at this link \github.}
  % conclusions heading (optional), leave it empty if necessary 
   {}
   \keywords{methods: statistical – gravitational lensing: weak – cosmology: large-scale structure of Universe
               }

@@

%-------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{Introduction}
Weak gravitational lensing by the large-scale structure is caused by the presence of foreground matter bending the light emitted by background galaxies. Being sensitive to the large-scale structure of the universe, it is one of the most promising tools for investigating the nature of dark energy, the origin of the accelerating expansion of the universe and estimating cosmological parameters. Future cosmological surveys, such as the Legacy Survey of Space and Time (LSST) of the Vera C. Rubin Observatory \citep{ivezic2019lsst}, the Nancy Grace Roman Space Telescope \citep{spergel2015wide}, and  the Euclid Mission \citep{laureijs2011euclid}, will rely on weak gravitational lensing as one of the principal physical probes to address unresolved questions in current cosmology.
 As these surveys become deeper, they will be able to access more non-Gaussian features of the matter fields. This makes the standard weak lensing analyses, which rely on two-point statistics such as the 2-point shear correlation or the angular power spectrum, sub-optimal. These analyses are unable to fully capture the non-Gaussian information imprinted in the lensing signal and can only access Gaussian information.

@@

In this paper, we present a comparison of the performance of various neural network-based compression schemes within the Likelihood-Free Inference (LFI) framework. These methods differ in terms of the loss functions used to train the neural network, but they share the same inference strategy based on neural density estimation. To address the need for result validation in the LFI context, we employ a Bayesian forward modeling approach that enables inference on the joint posterior of the cosmological parameters and the convergence field. The forward model is based on \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\url{SbiLens}}, a Python package designed for weak lensing inference with differentiable simulator implemented in Jax. \texttt{SbiLens} allows the sampling of convergence maps in a tomographic setting, accounting for the cross-correlation between different redshift bins.

The paper is structured as follows: in \autoref{Sec:Motivation}, we illustrate the motivation behind this work. In \autoref{Sec:the SBILens framework}, we introduce the \texttt{SbiLens} framework and describe the simulated data used in this work. In \autoref{Sec:experiment}, we detail the inference strategy and the three different approaches we used: the power spectrum, the map-based inference based on the Bayesian forward-modeling, and the map-based inference based on the LFI. In the same section, we also provide a detailed overview of different neural compression strategies. In \autoref{Sec:results}, we discuss the results and validate the LFI approaches. Finally, we conclude in \autoref{Sec:conclusion}. 
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{Motivation}\label{Sec:Motivation}
%--------------------------------------------------------------------
With the increased statistical power of stage IV surveys, our cosmological analysis should not rely on the measurement of sub-optimal summary statistics, that may not fully capture the non-Gaussian information present in the lensing field at the scales accessible to future surveys. In this paper we introduce a forward model that directly extracts information from the raw pixel data, rather than relying on the analytical evaluation of summary statistics. By doing so, we aim to preserve all available information and facilitate the incorporation of systematic effects and the combination of multiple cosmological probes through joint simulations.
In this context, the simulator of the observables serves as our physical model, where each component is tractable. These models, often refereed to as \textit{probabilistic program}, can be described as follows. They take as input a vector parameter $\bm{\theta}$, which describes the underlying deterministic model. Then, they sample internal states, dubbed \textit{latent variables}, from the distribution $\bm{z}_i \sim p_i(\bm{z}_i|\bm \theta, \bm{z}_{<i})$, that can be directly or indirectly related to a physically meaningful state of the system. Finally, they generate the output $\bm x$ from the distribution $\ p(\bm d|\bm \theta)$, where $\bm{d}$ represents the observations.

The ultimate goal of Bayesian inference in cosmology is to compute the posterior distribution:
\begin{equation}\label{Eq:posterior}
     p(\bm{\theta}|\bm{x})= 
     \frac{p(\bm{x}|\bm{\theta})p(\bm{\theta})}
     {\int d\bm{\theta'}p(\bm{x}|\bm{\theta}')p(\bm{\theta}')},
\end{equation}
however, a problems arises because the marginal likelihood $p(\bm{x}|\bm{\theta})$ is typically intractable:
\begin{equation}
    p(\bm{x}|\bm{\theta})=\int p(\bm{x},\bm{z}|\bm{\theta}) d\bm{z}=\int p(\bm{x}|\bm{z},\bm{\theta})p(\bm{z}|\bm{\theta}) d\bm{z}.
\end{equation}
as it corresponds to an integral over all possible trajectories throughout the latent space.
To overcome this limitation while still capturing the full information content of the data, two different approaches have been proposed in the literature. Although these approaches are often referred to by different names, in this work, we will make the following distinction:
%--------------------------------------------------------------------
\paragraph{\textbf{Explicit inference}} referring to all Likelihood-based inference approaches.  
This approach involves treating the simulator as a probabilistic model and performing inference over the joint posterior:
\begin{equation}
        p(\bm{\theta},\bm{z}|\bm{x})\propto  p(\bm{x},\bm{z}|\bm{\theta}) p(\bm{z}|\bm{\theta}).
\end{equation}
In the context of the forward model and full field analysis, fall in this category the Bayesian Hierarchical Models (BHMs).
A Bayesian forward approach involves using a given physical model to predict observations and then comparing these predictions with observations to infer the parameters of the model.
However, performing a global analysis on complex and large datasets, such as a lensing dataset, is not feasible. Therefore, the idea of dividing the global analysis into several subgroups, analyzing them in multiple steps, and using the output of each of those as input for the next one. From here, the \textit{hierarchical} nature of this methodology.
%--------------------------------------------------------------------
\paragraph{\textbf{Implicit inference}} referring to all the approaches not relying on an analytical model to describe the signal, but rather on learning the likelihood from simulations. This second class of approaches involves treating the simulator as a black box with only the ability to sample from the joint distribution:
\begin{equation}
    (\bm{x}, \bm{\theta})\sim p(\bm{x}, \bm{\theta}).
\end{equation}
Within this class of methods, we can differentiate between more traditional methods such as Approximate Bayesian Computation (ABC) and Density Estimation Likelihood-Free Inference (DELFI) methods. ABC employs rejection-criteria based approaches to approximate the likelihood by comparing simulations with data. 
In this work, our focus will be on DELFI methods, which approach the inference task as a density estimation problem. \\
We can divide Likelihood-Free Inference approaches into two separate steps:
\begin{enumerate}
    \item  Automatically learning an optimal low-dimensional summary statistic.
    \item Using Neural Density Estimation in low dimensions to infer the target distributions.
\end{enumerate}
In the first step, we introduce a parametric function $f_{\varphi}$ such that:
     \begin{equation}
         \bm{t}=F_{\varphi}(\bm{\theta}),
     \end{equation}
which aims to reduce the dimensionality of the data while preserving information. Typically, the compressed statistics $\bm{t}$ is assumed to have the same dimension of $\bm{\theta}$. \\
In the second step, Neural Density Estimation can target either building an estimate $p_{\phi}$ of the likelihood function $p(\bm{x}|\bm{\theta})$ (referred to as the Neural Likelihood Estimation (NLE) task), or targeting the posterior distribution $p(\bm{\theta}|\bm{x})$, (known as Neural Posterior Estimation (NPE) task).

@@

%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{The SBILens framework}\label{Sec:the SBILens framework}
\subsection{Lognormal Modeling}\label{Sec:Lognormal Modeling}
For various cosmological applications, the non-Gaussian field can be modeled as a Lognormal field \citep{coles1991lognormal,bohm2017bayesian}.
This model offers the advantage of generating the matter or convergence field rapidly while allowing the extraction of information beyond the two-point statistics. 
Although studies demonstrated that this model fails in describing the 3D field \citep{klypin2018density}, it properly describes the 2D convergence field \citep{clerkin2017testing, xavier2016improving}.
Assuming a simulated Gaussian convergence map $\kappa_g$, whose statistical properties are fully described by its power spectrum $C_{\ell}$ we know that this model is not a suitable representation of late-time and more evolved structures. One potential solution is to find a transformation $f(\kappa_g)$ of this map that captures the non-Gaussian features in the convergence field. In doing so, it is crucial to ensure that the transformed map maintains the correct mean and variance, effectively recovering the correct two-point statistics.
Denoting $\mu$ and $\sigma_g^2$ the mean and covariance matrix of $\kappa_g$ respectively, we can define the transformed convergence $\kappa_{ln}$ as a shifted lognormal random field:
\begin{equation}\label{Eq:log_norm_kappa}
    \kappa_{ln}=e^{\kappa_{g}}-\lambda, 
\end{equation}
where $\lambda$ is a free parameter that determines the shift of the lognormal distribution. The convergence $\kappa$ in a given redshift bin is fully determined by the shift parameter $\lambda$, the mean $\mu$ of the associated Gaussian field $\kappa_g$, and its variance $\sigma_{g}^2\equiv \xi_g$.
The correlation of the lognormal field, denoted as $\xi_{ln}$, is also a function of these variables and is related to $\xi^{ij}_g$ through the following equations:
\begin{align}
    \xi^{ij}_{ln}(\theta) & \equiv \lambda_i \lambda_j (e^{ \xi^{ij}_g(\theta)}-1) \nonumber \\ 
    \xi^{ij}_g(\theta)&=\log{\left[ \frac{\xi^{ij}_{ln}(\theta)}{\lambda_i \lambda_j}+1\right ]}. \label{Eq:log_norm_corr}
\end{align}
Here $i$ and $j$ define a pair of redshift bins.
The parameter $\lambda$, also known as \textit{minimum convergence
parameter}, defines the lowest values for all possible values of $\kappa$.
The modeling of the shift parameter can be approached in various ways. For example, it can be determined by matching moments of the distribution \citep{xavier2016improving} or by treating it as a free parameter \citep{hilbert2011cosmic}. In general, the value of $\lambda$ depends on the redshift, cosmology, and the scale of the field at which smoothing is applied.

@@

@@

% ############# PRIOR COSMO TABLE  #############

@@

For each redshift bin, we assume Gaussian noise with mean zero and variance given by:

@@

% ############# SURVEY SPECIFICATION TABLE  #############

@@

% ############# PLOT REDSHIFT DISTRIBUTION  #############

@@

@@

This means that this loss function is actually sub-optimal for two reasons:
\begin{itemize}
    \item The summary extracted by the neural network $t$ is, similarly to the MSE case only an estimate of the mean of 
    the posterior distribution, which is not guaranteed to be sufficient.
    \item Because of the Gaussian variational assumption, the variational posterior can potentially be biased with respect to the true posterior, and thus the mean of this Gaussian approximation may be biased with respect to the true posterior mean.
\end{itemize}
 A simple MSE would thus seem to be preferable.

@@

\paragraph{\textcolor{violet}
{Information Maximising Neural Networks (IMNNs)}}. \LM{TODO: condensed introduction; emphasize learning compression at single point; network learns score function which generalises smoothly away from fiducial \cite{Makinen_2021, Makinen_2022, Charnock_2018}, Wandelt (in prep.) 
\subsection{Compression strategy}
We aim to examine the impact of different loss functions used to train the compressor on the final cosmological constraints. We employ the same convolutional neural network architecture to build the compressor. Specifically, we utilize a deep ResNet architecture \citep{he2016deep}, the ResNet-18, implemented using Haiku \citep{haiku2020github}, a Python deep learning library built on top of JAX. While the different compression strategies share the same architecture, the training strategies for VMIM and GNLL differ significantly. To train the Neural compressor under VMIM and GNLL, we jointly optimize the weights of the Neural Network $F_{\varphi}$ (the compressor) and the parameters of the variational distribution $q_{\varphi'}$. For VMIM, the variational distribution $q(\bm{\theta}|\bm{t};\varphi')$ is modeled using a Normalizing Flow, while for GNLL, a Conditional Multivariate Gaussian estimator is employed. The Normalizing Flow used for the inference strategy is also adopted for the variational distribution required in the compression procedure. For the Multivariate Gaussian estimator, we implement a three-layer neural network. The first two layers consist of fully connected layers with a Leaky Rectified Linear Unit activation function and a hyperbolic tangent activation function, respectively. The number of channels is increased to 256 and 128 for the first two layers, respectively. The last layer is a fully connected layer without an activation function, responsible for mapping the output of the first two layers to the desired output size (mean and covariance of the Gaussian distribution).
\subsubsection{Inference strategy}\label{Sec:Inference_strategy}
By comparing the posteriors $p(\theta| t)$ obtained with different compression strategies, we can assess the sensitivity of the results to the choice of statistics. In particular, the more informative the summary statistics, the “closer” the posterior $p(\theta | t)$ is to the true posterior $p(\theta | x)$. In this section, we describe how we approximate the posterior distribution for each summary statistic t, enable us to use it as a fair benchmark for comparison (we describe the performance metric in section ??).\\
Since the likelihood $p(t|\theta)$ is intractable, we propose to do Neural Posterior Estimation \citep{npe1, npe2, npe3}, a Simulation-Based Inference (SBI) strategy that aims to directly approximate, from simulations, the posterior distribution 
    p(\theta | t) \propto \int p(t | \theta, z) p(z|\theta) p(\theta) dz\,.

@@

For each summary statistic, we approximate it posterior distribution using the same NF architecture, namely a RealNVP \citep{realnvp} with 4 coupling layers, shift and scale parameters are learned using a neural network with 2 layers of 128 neurons, activation functions are $silu$ \citep{silu}. And, for each summary statistic, the prior $p(\theta)$ used in the inference process remains the same.
\subsection{Result estimators}
To evaluate the results obtained from different approaches, we use two commonly employed estimators found in the literature. Their definitions are provided in the following sections.
\subsubsection{Figure of Merit (FoM)}
To quantify the constraining power of each inference strategy, we first employ the Figure of Merit (FoM), defined as follows:
\begin{equation}\label{Eq:Figure_of_merit}
    \text{FoM}_{\alpha \beta}=\sqrt{\text{det}(\Tilde{F}_{\alpha \beta})}.
Here, $\alpha$ and $\beta$ represent a pair of cosmological parameters, and $\Tilde{F}_{\alpha \beta}$ refers to the marginalized Fisher matrix. We calculate $\Tilde{F}_{\alpha \beta}$ as the inverse of the parameter space covariance matrix $C_{\alpha \beta}$, which is estimated from the HMC chains and from the Normalizing flow sampling. 
Under the assumption of a Gaussian covariance, the FoM defined in \autoref{Eq:Figure_of_merit} is proportional to the inverse of the $2-\sigma$ contours in the 2-dimensional marginalised parameter space of the $\alpha$ and $\beta$ pair.
\subsubsection{Kullback-Leibler divergence}
To measure the difference between the probability distribution obtained with different compressed summaries we compute the \textit{Kullback-Leibler divergence} or KL-divergence. As mentioned when we introduced the VMIM, the KL divergence is related to the relative entropy, and is a non-symmetric measure of the difference between two distributions. \denise{INCOMPLETE}
% ############# Begin Table BiblioSruvey  #############
        \begin{tabular}{ |p{3.5cm}|p{3.5cm}|p{3.5cm}|p{4cm}|  }
            Reference &  \makecell{Loss function} & \makecell{Inference \\ strategy} & \makecell{Output compressor}   \\
            \citet{2018PhRvD..97j3515G} & \makecell{MAE} &  \makecell{Likelihood \\ analysis} &\makecell{$\sigma_8, \Omega_m$}  \\
